\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{geometry}
\geometry{margin=1in}

\title{Independent Events in Two-Dice Experiments}
\author{}
\date{}

\begin{document}
\maketitle

\Large

Two events are said to be \emph{independent} if:
\[ P(A \cap B) = P(A)P(B)\]

Mutually exclusive events are never independent. In this section we look at dice rolling experiments which are both dependent and independent. 

The key point is that when events or \emph{not} independent the conditional probability doesn't equal the ``raw'' probability:
\[ P(B | A) \neq P(B) \]
In other words, knowing that $B$ occured gives you some predictive knowledge over whether $A$ will occur.
The more general formula is:
\[ P(A \cap B) = P(A)P(B|A)\]




\section*{Example 1: Sum Equals 7 and First Die is 6 [Independent]}

Consider rolling two fair six-sided dice. Define the events:

\[
A = \{\text{sum of the dice is 7}\} = \{(1,6),(2,5),(3,4),(4,3),(5,2),(6,1)\}
\]

\[
B = \{\text{first die shows 6}\} = \{(6,1),(6,2),(6,3),(6,4),(6,5),(6,6)\}
\]

\subsection*{Step 1: Probabilities}

\[
P(A) = \frac{6}{36} = \frac{1}{6}, \quad
P(B) = \frac{6}{36} = \frac{1}{6}
\]

\[
A \cap B = \{(6,1)\} \implies P(A \cap B) = \frac{1}{36}
\]

\subsection*{Step 2: Check Independence}

Two events are independent if:

\[
P(A \cap B) = P(A)P(B)
\]

Compute:

\[
P(A)P(B) = \frac{1}{6} \cdot \frac{1}{6} = \frac{1}{36} = P(A \cap B)
\]

Thus, \(A\) and \(B\) are \textbf{independent}.

\subsection*{Step 3: Intuition}

Knowing that the first die is 6 does not change the probability that the sum of the dice is 7.  
- Original probability: \(P(A) = 1/6\)  
- Conditional probability: \(P(A \mid B) = \frac{P(A \cap B)}{P(B)} = \frac{1/36}{6/36} = 1/6\)  

Since \(P(A \mid B) = P(A)\), the events are independent.

\bigskip

\section*{Example 2: Sum is Even and First Die is Even [Independent]}

Consider rolling two fair six-sided dice. Define the events:

\[
A = \{\text{sum of the dice is even}\}, \quad
B = \{\text{first die shows even}\}
\]

\subsection*{Step 1: Probabilities}

There are 36 equally likely outcomes. Half of them have an even sum:

\[
P(A) = \frac{18}{36} = 0.5
\]

Half of the outcomes have an even first die:

\[
P(B) = \frac{18}{36} = 0.5
\]

The intersection is outcomes where first die is even and sum is even:

\[
A \cap B = \{(2,2),(2,4),(2,6),(4,2),(4,4),(4,6),(6,2),(6,4),(6,6)\}
\]

\[
P(A \cap B) = \frac{9}{36} = 0.25
\]

\subsection*{Step 2: Check Independence}

\[
P(A)P(B) = 0.5 \cdot 0.5 = 0.25 = P(A \cap B)
\]

So \(A\) and \(B\) are \textbf{independent}.

\subsection*{Step 3: Intuition}

- Event \(A\) depends on the parity of the sum.  
- Event \(B\) fixes the parity of the first die.  
- Whether the sum is even depends only on whether the two dice have the same parity.  
- If the first die is even, the sum is even if the second die is even (50\% chance).  
- If the first die is odd, the sum is even if the second die is odd (50\% chance).  

In both cases, \(P(A \mid B) = P(A) = 0.5\).  
Knowing that the first die is even does not change the probability that the sum is even.

\section*{Example 3: Sum Even and Sum Greater than 7 [Dependent]}

Consider rolling two fair six-sided dice. Define the events:

\[
A = \{\text{sum of the dice is even}\}, \quad
B = \{\text{sum of the dice is greater than 7}\}
\]

\subsection*{Step 1: Probabilities}

- There are 36 equally likely outcomes in total.  
- Number of outcomes with an even sum: 18  
\[
P(A) = \frac{18}{36} = 0.5
\]

- Number of outcomes with sum greater than 7: 15  
\[
P(B) = \frac{15}{36} \approx 0.4167
\]

- Intersection (sum even \textbf{and} sum $> 7$): sums 8, 10, 12 → 9 outcomes  
\[
P(A \cap B) = \frac{9}{36} = 0.25
\]

\subsection*{Step 2: Check Independence}

Two events are independent if:

\[
P(A \cap B) = P(A)P(B)
\]

Compute:

\[
P(A)P(B) = 0.5 \cdot 0.4167 = 0.2083 \neq 0.25
\]

\textbf{Conclusion:} \(A\) and \(B\) are \textbf{not independent}.

\subsection*{Step 3: Conditional Probability and Intuition}

The conditional probability of \(A\) given \(B\) is:

\[
P(A \mid B) = \frac{P(A \cap B)}{P(B)} = \frac{0.25}{0.4167} \approx 0.6
\]

- Original probability of sum even: \(P(A) = 0.5\)  
- Probability of sum even given sum $>7$: \(P(A \mid B) \approx 0.6\)

\bigskip

\textbf{Intuition:}  

- Event \(A\) depends on the parity of the sum.  
- Event \(B\) restricts outcomes to sums greater than 7.  
- Among sums $>7$ (8, 9, 10, 11, 12), there are proportionally more even sums (8, 10, 12) than odd sums (9, 11).  
- Therefore, knowing that \(B\) occurred increases the probability of \(A\), which is exactly why \(A\) and \(B\) are \textbf{dependent}.


\section*{Example 4: At Least One 6 and Sum Greater Than 10 [Dependent]}

Consider rolling two fair six-sided dice. Define the events:

\[
C = \{\text{at least one die shows 6}\}, \quad
D = \{\text{sum of the dice is greater than 10}\}
\]

\subsection*{Probabilities}

- Total outcomes: 36  
- Outcomes with at least one 6: 11  
\[
P(C) = \frac{11}{36} \approx 0.3056
\]  

- Outcomes with sum $> 10$: 3  
\[
P(D) = \frac{3}{36} \approx 0.0833
\]  

- Intersection (at least one 6 and sum $>10$): 3 outcomes  
\[
P(C \cap D) = \frac{3}{36} = 0.0833
\]

\subsection*{Check Independence}

Two events are independent if \(P(C \cap D) = P(C) P(D)\).  

\[
P(C)P(D) = 0.3056 \cdot 0.0833 \approx 0.0254 \neq 0.0833
\]

\textbf{Conclusion:} \(C\) and \(D\) are \textbf{not independent}.

\subsection*{Conditional Probability and Intuition}

\[
P(D \mid C) = \frac{P(C \cap D)}{P(C)} = \frac{0.0833}{0.3056} \approx 0.2727
\]

- Original probability of sum $>10$: \(P(D) = 0.0833\)  
- Probability of sum $>10$ given at least one die is 6: \(P(D \mid C) \approx 0.2727\)

\textbf{Intuition:}  

- Event \(D\) requires a high sum ($>10$).  
- Event \(C\) ensures at least one die is 6.  
- High sums are more likely if at least one die is 6, so knowing \(C\) occurred significantly increases the probability of \(D\), showing dependence.




\end{document}

